tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-28 paper.tex
Pass through discussion

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using familiar Python data science tools.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals, for sequences of length 50, 100 and 200Mb.
For each sample size and sequence length,
we first simulate an ARG under a simple coalescent model with
human-like parameters using~\texttt{stdpopsim}~\citep{adrion2020}.
We then run a quantitative trait simulation
with 1000 causal sites on this ARG using \texttt{tstrait},
and report the mean CPU time over 10 independent replicates.
Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
modest for all these simulations.
To demonstrate its scalability we  applied \texttt{tstrait} to the
French-Canadian simulations~\cite{anderson2023} of chromosome 9
discussed in the introduction.
% TODO redo this for 100 causal sites @daiki.
It took 6 seconds and XXX GB of RAM to
simulate a trait with 100 causal sites using \texttt{tstrait}.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for the ARG estimated from chromosome 9 of the 1000 Genomes
project data~\citep{kelleher2019} which has YYY samples
and ZZZ variant sites. This took 0.X seconds and Y GB of RAM.

\section{Conclusion}
[WIP]
There is a pressing need for large-scale truth datasets to support
GWAS method development. One approach, exemplified by
\citet{wharrie2023hapnest}, is to generate synthethic genetic variation
based on publicly available data and to make these available. While
there are certainly advantages to this method
publicly

For example, the \texttt{plink} formatted data file of chromosome 9
for 1 million individuals provided for download is 78GB.

% We believe that in the coming years, simulation in tree sequence encoding will
% attract more users in need of conducting genetic simulations with complex
% population structure. Multiple studies highlight the advantages of using the
% ARG data structure in GWAS \citep{link2023tree,nowbandegani2023extremely,zhang2023}.
% ARG data
% structures are becoming increasingly useful in genetic studies, as it is possible
% to accurately infer biobank-scale genealogies from sequencing data
% \citep{zhang2023}, and simulate large sample whole-genome sequences with
% spatiotemporal metadata \citep{anderson2023}. These advances in ARG presents a
% real opportunity for people to further understand the effects of complex
% population structure on quantitative traits. The \texttt{tstrait} package is an
% essential piece of infrastructure that lets people explore these possibilities.

>>>>

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG toolkit~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
Simulating a phenotype for an ARG with default parameter
values requires only a few lines of code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python,aboveskip=1em,belowskip=1em]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The user can either specify a number of causal sites to be chosen randomly
along the genome (one, by default), or can directly provide the causal
sites as input. Combined with the detailed information about
mutations recorded in a \texttt{tskit} ARG, explicitly specifying causal
sites allows us to model many different types of trait, for example
those associated with mutations arising in a particular population
or time interval.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed using familiar Python data science tools.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales.
Supplementary Fig~\ref{fig:time}A shows how trait simulation time scales
with number of individuals on human-like coalescent simulations
generated using~\texttt{stdpopsim}~\citep{adrion2020}.
For each simulated ARG, we simulate a trait with 1000 causal sites
and report the mean CPU time over 10 independent replicates.
Supplementary Fig~\ref{fig:time}B shows that RAM usage is very
modest. To emphasise scalability, we also
applied \texttt{tstrait} to the simulations of 1.4 million
French-Canadians discussed in the introduction.
% TODO redo this for 100 causal sites @daiki.
It took 6 seconds and XXX GB of RAM to
simulate a trait with 100 causal sites.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for an ARG estimated from 1000 Genomes project
data~\citep{kelleher2019} which has YYY samples
and ZZZ variant sites. This took 0.X seconds and Y GB of RAM.

\section{Conclusion}
There is substantial interest in using inferred ARGs to improve
association testing
methods~\citep{zhang2023,link2023tree,nowbandegani2023extremely},
and there is a pressing need for a well-tested, efficient
and user-friendly means of simulating phenotypes on ARGs.
Highly realistic simulations conditioned on large
pedigrees~\citep{anderson2023} provide an exciting opportunity
to test the effects of intricate population structure on
GWAS, and we hope that \texttt{tstrait} will facilitate these
investigations. \texttt{Tstrait}'s modular architecture
and flexible specification of causal sites
should provide the opportunity to explore new avenues of research,
and an extensible platform for future development.
